---
title: "Niche Construction, larval and maternal, between populations of *Onthophagus taurus*"
output:
  pdf_document: default
  html_document: default
---

##Loading the required packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here) #For relative file-paths

#Packages used for statistical analysis
library(lme4) #For linear mixed-effects models
library(lmerTest) #To perform statistical tests on the objects created with lme4
library(dplyr) #To filter data
library(blmeco) #To do an overdispersion test
library(alr4) #For the ANOVA function
library(arm) #To get SE of random effects

#Packages used for figures
library(visreg)
library(effects)
library(ggplot2)
```

## Importing the data
```{r cars}
mydata = read.csv("OnthophagusNCdata.csv", header = T, na.strings=c("","NA"))
```

#Making sure that the numeric variables and factors are stored as such:
```{r}
mydata$Mass.3rd.Ins <- as.numeric(as.character(mydata$Mass.3rd.Ins))
mydata$Mass.Pupa <- as.numeric(as.character(mydata$Mass.Pupa))
mydata$Dung.batch <- as.factor(as.character(mydata$Dung.batch))
mydata$Sample.number <- as.factor(as.character(mydata$Sample.number))
mydata$Batch <- as.factor(as.character(mydata$Batch))
```

#Creating a log-pupa mass column
```{r}
mydata$logMass.Pupa <- log(mydata$Mass.Pupa*1000)
mydata$Mass.Pupa <- mydata$Mass.Pupa*1000
```

#Creating a column of factors to use in making plots
```{r}
mydata$factors <- interaction(mydata$NC, mydata$Temp, mydata$Pop) #For plots
```

#Creating a for binary survival data
```{r}
#Making a column with binary Survival data
mydata$Survival <- ifelse(is.na(mydata$Day.Dead),1,0)
mydata$Survival <- as.factor(mydata$Survival)
```

#Optional code to include a suspicious datapoint
Individual 120 seems to have reached pupation in only 10 days, this is highly suspicious and likely a transcribing error. Therefore that data-point was removed from the analyses. If you wish to include this data-point in the analysis, remove the octothorp ("#") from the beginning of both lines and run this block of code.
```{r}
#mydata$Days.to.Pupa[120] <- 10
#mydata$Date.Pupa[120] <- "2017-07-24"
```

## Filter or subset data
Since some individuals were injured, of which some survived and some died, but that data prior to the injury is useful, I subset data for it to contain only uninjured individuals (Survival, Mass3rd, logMassPupa, etc.). The other subsets (SurvivalwMaybe, Mass3rdwMaybe, logMassPupawMaybe, etc.) contain individuals that were injured but survived.
```{r}
Survival <- subset(mydata, mydata$Exclude.death == "No")
SurvivalwMaybe <- subset(mydata, subset = mydata$Exclude.death %in% c("No","Maybe"))

Mass3rd <- subset(mydata, mydata$Exclude.Mass.3rd == "No")
Mass3rdwMaybe <- subset(mydata, subset = mydata$Exclude.Mass.3rd %in% c("No","Maybe"))

logMassPupa <- subset(mydata, mydata$Exclude.Mass.pupa == "No")
logMassPupawMaybe <- subset(mydata, subset = mydata$Exclude.Mass.pupa %in% c("No","Maybe"))

DaystoPupa <- subset(mydata, mydata$Exclude.date.3rd=="No")
DaystoPupawMaybe <- subset(mydata, subset = mydata$Exclude.date.3rd %in% c("No","Maybe"))

GramstoL3 <- subset(mydata, mydata$Exclude.date.3rd == "No")
GramstoL3 <- subset(GramstoL3, mydata$Exclude.Mass.3rd=="No")
GramstoL3wMaybe <- subset(mydata, subset = mydata$Exclude.date.3rd %in% c("No","Maybe"))
GramstoL3wMaybe <- subset(mydata, subset = mydata$Exclude.Mass.3rd %in% c("No","Maybe"))

AdultSize <- subset(mydata, mydata$Exclude.adult == "No")
AdultSizewMaybe <- subset(mydata, subset = mydata$Exclude.adult %in% c("No","Maybe"))
```

## Analysis of Mass of third instar
```{r}
#Tests whether 'mass of 3rd instar' are normally distributed
shapiro.test(Mass3rd$Mass.3rd.Ins) 

#Linear mixed models comparing larval mass at day 10 of third instar, this is reported in Table S3
MTModel1 = lmer(Mass.3rd.Ins*1000 ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
              + Pop:NC:Temp + (1|Batch), data = Mass3rd)
summary(MTModel1)
Anova(MTModel1)

#Linear mixed models comparing larval mass at day 10 of third instar, this is reported in Table S3
MTModel2 = lmer(Mass.3rd.Ins*1000 ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
              + (1|Batch), data = Mass3rd)
summary(MTModel2)
Anova(MTModel2)

#Linear mixed models comparing larval mass at day 10 of third instar, this is reported in Table 1
MTModel3 = lmer(Mass.3rd.Ins*1000 ~ Pop + NC + Temp
              + (1|Batch), data = Mass3rd)
summary(MTModel3)
Anova(MTModel3)


#The following three models are the same, but include individuals that survived injury
#These models are not reported in the paper.
MTModel1wM = lmer(Mass.3rd.Ins*1000 ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
              + Pop:NC:Temp + (1|Batch), data = Mass3rdwMaybe)
summary(MTModel1wM)
Anova(MTModel1wM)

MTModel2wM = lmer(Mass.3rd.Ins*1000 ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
              + (1|Batch), data = Mass3rdwMaybe)
summary(MTModel2wM)
Anova(MTModel2wM)

MTModel3wM = lmer(Mass.3rd.Ins*1000 ~ Pop + NC + Temp
              + (1|Batch), data = Mass3rdwMaybe)
summary(MTModel3wM)
Anova(MTModel3wM)
```

##Figures of Third instar mass (used to create Figure S1)
```{r}
#To visualize the (significant) effect of larval niche construction
#This was used to create panel B of Figure S1
Mass3rdNCFig <- visreg(MTModel3, "NC", gg = FALSE, asp=0.005)
Mass3rdNCFig
#svg("Mass3rdNCFig.svg")
#plot(Mass3rdNCFig)
#dev.off()

#To visualize the (significant) effect of maternal niche construction
#This was used to create panel A of Figure S1
Mass3rdTempFig <- visreg(MTModel3, "Temp", gg = FALSE, asp=0.005)
Mass3rdTempFig
#svg("Mass3rdTempFig.svg")
#plot(Mass3rdTempFig)
#dev.off()

#To visualize the (non-significant) effect of population
#Because this was non-significant, this figure was not used
Mass3rdPopFig <- visreg(MTModel3, "Pop", gg = FALSE, asp=0.005)
Mass3rdPopFig
```

##Analysis of days to pupa
```{r}
#Tests whether the number of days to pupation are normally distributed
shapiro.test(as.numeric(DaystoPupawMaybe$Days.to.Pupa))

#Quantile-Quantile Plot
qqnorm(as.numeric(DaystoPupawMaybe$Days.to.Pupa))
qqline(as.numeric(DaystoPupawMaybe$Days.to.Pupa))

#Generalized linear mixed models for time to pupation (results are reported in Table 2)
DPModel1 = glmer(as.numeric(Days.to.Pupa) ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp + Pop:NC:Temp + (1|Batch), data = DaystoPupa, family = poisson)
dispersion_glmer(DPModel1) #to test for overdispersion, should not be above 1.4
summary(DPModel1)
Anova(DPModel1)

#Generalized linear mixed models for time to pupation, including injured individuals that survived their injury
DPModel1wM = glmer(as.numeric(Days.to.Pupa) ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp + Pop:NC:Temp + (1|Batch), data = DaystoPupawMaybe, family = poisson)
dispersion_glmer(DPModel1wM) #to test for overdispersion, should not be above 1.4
summary(DPModel1wM)
Anova(DPModel1wM)
```

##Figures of days to pupa using the package effects
```{r}
#Extracting the data to be plotted from the model
efTime <- effect("Pop:NC:Temp", DPModel1)
#summary(efTime)

#Putting that extracted data into a dataframe
xTime <- as.data.frame(efTime)
#xTime

#This is the plot for days to pupation, used to create Figure 3
#(The two commented-out lines can be used to export the svg used as the basis for Figure 3).
#svg("NCDaysToPupa.svg")
ggplot(xTime, aes(Temp, fit, color=Pop, shape=NC)) + geom_point() + geom_errorbar(aes(ymin=fit-se, ymax=fit+se), width=0.2) + coord_fixed(ratio=0.15) + theme_bw(base_size=12)
#dev.off()
```

## Analysis of log Mass of pupae
```{r}
#Test for normality of mass of pupa shows it is not normally distributed
shapiro.test(logMassPupawMaybe$Mass.Pupa)
#Test for normality of log(mass of pupa) shows it is normally distributed
shapiro.test(logMassPupawMaybe$logMass.Pupa)

#Linear mixed model comparing log(pupa mass), this is reported in Table S4
LPModel1 = lmer(logMass.Pupa ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
              + Pop:NC:Temp + (1|Batch), data = logMassPupa)
summary(LPModel1)
Anova(LPModel1)

#Linear mixed model comparing log(pupa mass), this is reported in Table S4
LPModel2 = lmer(logMass.Pupa ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
              + (1|Batch), data = logMassPupa)
summary(LPModel2)
Anova(LPModel2)

#Linear mixed model comparing log(pupa mass), this is the model reported in Table 1 for this analysis
LPModel3 = lmer(logMass.Pupa ~ Pop + NC + Temp
              + (1|Batch), data = logMassPupa)
summary(LPModel3)
Anova(LPModel3)

#The following three models are the same, except that they include individuals that survived injury
#These models are not reported in the paper.
LPModel1wM = lmer(logMass.Pupa ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
              + Pop:NC:Temp + (1|Batch), data = logMassPupawMaybe)
summary(LPModel1wM)
Anova(LPModel1wM)

LPModel2wM = lmer(logMass.Pupa ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
              + (1|Batch), data = logMassPupawMaybe)
summary(LPModel2wM)
Anova(LPModel2wM)

LPModel3wM = lmer(logMass.Pupa ~ Pop + NC + Temp
              + (1|Batch), data = logMassPupawMaybe)
summary(LPModel3wM)
Anova(LPModel3wM)
```


##Figure of log(pupa mass)
```{r}
#This is the code used to create panel B of figure S2
#To export a copy of the vector file used to create the figure, run the three lines that are commented out
MassPupaNCFig <- visreg(LPModel3, "NC", trans=exp, partial=TRUE, gg = FALSE)
MassPupaNCFig
#svg("MassPupaNCFig.svg")
#visreg(LPModel2, "NC", trans=exp, partial=TRUE, gg = FALSE)
#dev.off()

#This is the code used to create panel A of figure S2
#To export a copy of the vector file used to create the figure, run the three lines that are commented out
MassPupaTempFig <- visreg(LPModel3, "Temp", trans=exp, partial=TRUE, gg = FALSE)
MassPupaTempFig
#svg("MassPupaTempFig.svg")
#visreg(LPModel2, "Temp", trans=exp, partial=TRUE, gg = FALSE)
#dev.off()
```


## Analysis of survival
```{r}
#Generalized linear mixed models for survival (results are reported in Table S6)
SModel1 = glmer(as.factor(Survival) ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp + Pop:NC:Temp + (1|Batch), data = Survival, family = binomial, control=glmerControl(optimizer = "bobyqa"))
summary(SModel1)
Anova(SModel1)

#Generalized linear mixed models for survival (results are reported in Table 2)
SModel2 = glmer(Survival ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp + (1|Batch), data = Survival, family = binomial, control=glmerControl(optimizer = "bobyqa"))
summary(SModel2)
Anova(SModel2)

#Generalized linear mixed models for survival, including injured individuals that survived their injury
#These are not reported in the paper
SModel1wM = glmer(as.factor(Survival) ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp + Pop:NC:Temp + (1|Batch), data = SurvivalwMaybe, family = binomial, control=glmerControl(optimizer = "bobyqa"))
summary(SModel1wM)
Anova(SModel1wM)

SModel2wM = glmer(Survival ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp + (1|Batch), data = SurvivalwMaybe, family = binomial, control=glmerControl(optimizer = "bobyqa"))
summary(SModel2wM)
Anova(SModel2wM)

#Code to get a summary of sample size across populations for each treatment
#Making two-way factors column for summary 
Survival$twowayfactors <- interaction(Survival$NC, Survival$Temp) #For summary
SurvivalwMaybe$twowayfactors <- interaction(SurvivalwMaybe$NC, SurvivalwMaybe$Temp) #For summary

#Using dplyr to get a summary
Survival %>%
  group_by(twowayfactors, Survival) %>%
  summarise (n = n())

```

##Figure of survival (with package effects), used to make Figure 2
```{r}
#Extracting the effects of maternal and larval niche construction from the generalized linear mixed models for survival 
effectMortality <- effect("NC:Temp", SModel2)
#summary(ef)
#Putting the just-extracted data into a data-frame
effectMortalityDf <- as.data.frame(effectMortality)
#Adding a column for mortality (i.e., 1 - survival)
effectMortalityDf$Mort <- 1 - effectMortalityDf$fit
effectMortalityDf

#Mortality plot (with standard error error-bar above, but not below), this was used as the basis for Figure 2
MortTempNCDynamite <- ggplot(effectMortalityDf, aes(NC, Mort, color=Temp, fill=Temp)) + geom_bar(stat="identity", position="dodge") + geom_errorbar(aes(ymin=Mort, ymax=Mort+se), width=0.4, position=position_dodge(width=0.9)) + theme_bw(base_size=12) + coord_fixed(ratio=1.55)
MortTempNCDynamite

#Run the code below to export a vector format of the figure used as the basis for Figure 2
#svg("MortTempNCDynamite.svg")
#MortTempNCDynamite <- ggplot(x, aes(NC, Mort, color=Temp, fill=Temp)) + geom_bar(stat="identity", position="dodge") + geom_errorbar(aes(ymin=Mort, ymax=Mort+se), width=0.4, position=position_dodge(width=0.9)) + theme_bw(base_size=12) + coord_fixed(ratio=1.55)
#MortTempNCDynamite
#dev.off()
```



## Analysis of adult size
```{r}
#Test for normality of adult size, shows it is normally distributed
shapiro.test(AdultSize$Adult.Body.Size)

#Linear mixed model comparing adult size, this is reported in Table S5
ABSModel1 = lmer(Adult.Body.Size ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
              + Pop:NC:Temp + (1|Batch), data = AdultSize)
summary(ABSModel1)
Anova(ABSModel1)

#Linear mixed model comparing adult size, this is reported in Table S5
ABSModel2 = lmer(Adult.Body.Size ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
            + (1|Batch), data = AdultSize)
summary(ABSModel2)
Anova(ABSModel2)

#Linear mixed model comparing adult size, this is the model reported in Table 1 for this analysis
ABSModel3 = lmer(Adult.Body.Size ~ Pop + NC + Temp + (1|Batch), data = AdultSize)
summary(ABSModel3)
Anova(ABSModel3)

#The following three models are the same, except that they include individuals that survived injury
#Test for normality of adult size (including individuals that survived injury), shows it is normally distributed
shapiro.test(AdultSizewMaybe$Adult.Body.Size)
#These models are not reported in the paper.
ABSModel1wM = lmer(Adult.Body.Size ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
              + Pop:NC:Temp + (1|Batch), data = AdultSizewMaybe)
summary(ABSModel1wM)
Anova(ABSModel1wM)

ABSModel2wM = lmer(Adult.Body.Size ~ Pop + NC + Temp + Pop:NC + Pop:Temp + NC:Temp 
            + (1|Batch), data = AdultSizewMaybe)
summary(ABSModel2wM)
Anova(ABSModel2wM)

ABSModel3wM = lmer(Adult.Body.Size ~ Pop + NC + Temp + (1|Batch), data = AdultSizewMaybe)
summary(ABSModel3wM)
Anova(ABSModel3wM)
```

##Figures of adult size
```{r}
#This code was used to generate panel B of Figure 1
#To export a copy of the vector file used to create the figure, run the three lines that are commented out
SizeNCFig <- visreg(ABSModel3, "NC", gg = FALSE)
SizeNCFig
#svg("AdultSizeNCFig.svg")
#plot(SizeNCFig)
#dev.off()

#This code was used to generate panel A of Figure 1
#To export a copy of the vector file used to create the figure, run the three lines that are commented out
SizeTempFig <- visreg(ABSModel3, "Temp", gg = FALSE)
SizeTempFig
#svg("AdultSizeTempFig.svg")
#plot(SizeTempFig)
#dev.off()
```


##Counting number of injured individuals that survived injury
```{r}
mydata$Maybe.Mass.3rd <- ifelse(mydata$Exclude.Mass.3rd=="Maybe",1,0)
mydata$Maybe.Date.3rd <- ifelse(mydata$Exclude.date.3rd=="Maybe",1,0)
mydata$Maybe.Pupa <- ifelse(mydata$Exclude.Mass.pupa=="Maybe",1,0)
mydata$Maybe.all <- ifelse(mydata$Exclude.all=="Maybe",1,0)
mydata$Maybe.adult <- ifelse(mydata$Exclude.adult=="Maybe",1,0)
mydata$Sum.Maybe <- mydata$Maybe.Mass.3rd+mydata$Maybe.Date.3rd +mydata$Maybe.Pupa+mydata$Maybe.all+mydata$Maybe.adult #+mydata$Sum.Maybe
mydata$Injured.count <- ifelse(mydata$Sum.Maybe==0,0,1)
sum(mydata$Injured.count)
```

##Getting medians, sample sizes and standard errors to populate Table S2
```{r}
#Function to get mean, sample size, median and standard error for each metric
myFun <- function(x) {
  c(
    mean = round(mean(x, na.rm = TRUE),2), 
    median = round(median(x, na.rm = TRUE),2), 
    se = round(sd(x, na.rm = TRUE)/sqrt(length(which(!is.na(x)))),2),
    n=round(length(which(!is.na(x))),0)
    )
}

print("Mean, sample size, median and standard error for the mass of larvae 10 days into their third instar")
tapply(Mass3rd$Mass.3rd.Ins*1000, Mass3rd$factors, myFun)

print("Mean, sample size, median and standard error for the log(mass of pupae)")
tapply(logMassPupa$Mass.Pupa, logMassPupa$factors, myFun)

print("Mean, sample size, median and standard error for the number of days to pupation")
tapply(DaystoPupa$Days.to.Pupa, DaystoPupa$factors, myFun)

print("Mean, sample size, median and standard error for the adult body size")
tapply(AdultSize$Adult.Body.Size, AdultSize$factors, myFun)


#These are the same numbers but including individuals that survived injury
print("Mean, sample size, median and standard error for the mass of larvae 10 days into their third instar, including individuals that survived injury")
tapply(Mass3rdwMaybe$Mass.3rd.Ins*1000, Mass3rdwMaybe$factors, myFun)

print("Mean, sample size, median and standard error for the log(mass of pupae), including individuals that survived injury")
tapply(logMassPupawMaybe$Mass.Pupa, logMassPupawMaybe$factors, myFun)

print("Mean, sample size, median and standard error for the number of days to pupation, including individuals that survived injury")
tapply(DaystoPupawMaybe$Days.to.Pupa, DaystoPupawMaybe$factors, myFun)

print("Mean, sample size, median and standard error for the adult body size, including individuals that survived injury")
tapply(AdultSizewMaybe$Adult.Body.Size, AdultSizewMaybe$factors, myFun)
```
##Getting medians, sample sizes and standard errors to populate Table S1
```{r}
print("Number of individuals that survived to adulthood (and didn't)")
table(Survival$Survival, Survival$factors)

print("Percentages of individuals that survived to adulthood (and didn't)")
100*prop.table( table(Survival$Survival, Survival$factors) , 2)


#These are the same computations but including individuals that survived injury
print("Number of individuals that survived to adulthood (and didn't), including individuals that survived injury")
table(SurvivalwMaybe$Survival, SurvivalwMaybe$factors)

print("Percentages of individuals that survived to adulthood (and didn't), including individuals that survived injury")
100*prop.table( table(SurvivalwMaybe$Survival, SurvivalwMaybe$factors) , 2)

print("Total number of samples in each treatment, including individuals that survived injury")
#Some of these numbers are reported in the sampling intentions
summary(mydata$factors)
```
